---
title: "Google Data Analytics Case Study"
author: "Divit Jawa"
date: "7/5/2021"
output: prettydoc::html_pretty
vignette: > 
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

***
This analysis is for case study 1 from the Google Data Analytics Certificate (Cyclistic).  It’s originally based on the case study "'Sophisticated, Clear, and Polished’: Divvy and Data Visualization" written by Kevin Hartman (found here: https://artscience.blog/home/divvy-dataviz-case-study). We will be using the Divvy dataset for the case study.

# About the Company: 
In 2016, Cyclistic launched a successful bike-share offering. Since then, the program has grown to a fleet of 5,824 bicycles that are geotracked and locked into a network of 692 stations across Chicago. The bikes can be unlocked from one station and returned to any other station in the system anytime.

Until now, Cyclistic’s marketing strategy relied on building general awareness and appealing to broad consumer segments. One approach that helped make these things possible was the flexibility of its pricing plans: single-ride passes, full-day passes, and annual memberships. Customers who purchase single-ride or full-day passes are referred to as casual riders. Customers who purchase annual memberships are Cyclistic members.

Cyclistic’s finance analysts have concluded that annual members are much more profitable than casual riders. Although the pricing flexibility helps Cyclistic attract more customers, Moreno believes that maximizing the number of annual members will be key to future growth. Rather than creating a marketing campaign that targets all-new customers, Moreno believes there is a very good chance to convert casual riders into members. She notes that casual riders are already aware of the Cyclistic program and have chosen Cyclistic for their mobility needs.

Moreno has set a clear goal: Design marketing strategies aimed at converting casual riders into annual members. In order to do that, however, the marketing analyst team needs to better understand how annual members and casual riders dier, why casual riders would buy a membership, and how digital media could aect their marketing tactics. Moreno and her team are interested in analyzing the Cyclistic historical bike trip data to identify trends.

# Business Task: How do annual members and casual riders use Cyclistic bikes differently?

The key stakeholders are the Cyclistic company's marketing department, the company itself, and the Cyclistic company's casual usage bike-riders.

# Prepare Data

***

We will now install the required packages, load the data, and discuss any inconsistencies/discrepancies that need to be fixed.

```{r Install Packages}
#install.packages("tidyverse")
#install.packages("lubridate")

library(tidyverse)  #helps wrangle data
library(lubridate)  #helps wrangle date attributes
library(ggplot2)  #helps visualize data
```

### Read the Data

Data source: https://divvy-tripdata.s3.amazonaws.com/index.html under this [license](https://www.divvybikes.com/data-license-agreement).

```{r read datasets}
q2_2019 <- read_csv("csv/Divvy_Trips_2019_Q2.csv")
q3_2019 <- read_csv("csv/Divvy_Trips_2019_Q3.csv")
q4_2019 <- read_csv("csv/Divvy_Trips_2019_Q4.csv")
q1_2020 <- read_csv("csv/Divvy_Trips_2020_Q1.csv")
```

Let's check out the column names.

```{r colnames}
colnames(q3_2019)
colnames(q4_2019)
colnames(q2_2019)
colnames(q1_2020)

```

As we can see, the column names don't match up, so we'll rename them to match the q1_2020 data set since that's the latest one.

```{r rename cols}
(q4_2019 <- rename(q4_2019
                   ,ride_id = trip_id
                   ,rideable_type = bikeid 
                   ,started_at = start_time  
                   ,ended_at = end_time  
                   ,start_station_name = from_station_name 
                   ,start_station_id = from_station_id 
                   ,end_station_name = to_station_name 
                   ,end_station_id = to_station_id 
                   ,member_casual = usertype))

(q3_2019 <- rename(q3_2019
                   ,ride_id = trip_id
                   ,rideable_type = bikeid 
                   ,started_at = start_time  
                   ,ended_at = end_time  
                   ,start_station_name = from_station_name 
                   ,start_station_id = from_station_id 
                   ,end_station_name = to_station_name 
                   ,end_station_id = to_station_id 
                   ,member_casual = usertype))

(q2_2019 <- rename(q2_2019
                   ,ride_id = "01 - Rental Details Rental ID"
                   ,rideable_type = "01 - Rental Details Bike ID" 
                   ,started_at = "01 - Rental Details Local Start Time"  
                   ,ended_at = "01 - Rental Details Local End Time"  
                   ,start_station_name = "03 - Rental Start Station Name" 
                   ,start_station_id = "03 - Rental Start Station ID"
                   ,end_station_name = "02 - Rental End Station Name" 
                   ,end_station_id = "02 - Rental End Station ID"
                   ,member_casual = "User Type"))
```

We will now examine the structure of each of the dataframes.

```{r examine structure}
str(q1_2020)
str(q4_2019)
str(q3_2019)
str(q2_2019)
```
There are some discrepancies in the data type of the columns such as 'ride id' & 'rideable_type'. Let's convert those columns to match the data type for q1_2020.

```{r conver col-type}
q4_2019 <-  mutate(q4_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type))

q3_2019 <-  mutate(q3_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type))

q2_2019 <-  mutate(q2_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type))
```

Let's combine all the dataframes into one using *bind_rows()* method.

```{r combine dfs}
all_trips <- bind_rows(q2_2019, q3_2019, q4_2019, q1_2020)
```

There are certain fields that were not present in the 2020 data set, but were present in the 2019 data sets. So, we want to do a form of inner join, hence we will remove some columns.

```{r inner join}
all_trips <- all_trips %>%  
  select(-c(start_lat, start_lng, end_lat, end_lng, birthyear, gender, "01 - Rental Details Duration In Seconds Uncapped", "05 - Member Details Member Birthday Year", "Member Gender", "tripduration"))
```

Let's get some basic information about our newly cleaned and combined data set before we proceed.

```{r summary of df}
colnames(all_trips)  #List of column names
dim(all_trips)  #Dimensions of the data frame
head(all_trips)  #See the first 6 rows of data frame.
str(all_trips)  #See list of columns and data types (numeric, character, etc)
summary(all_trips)  #Statistical summary of data. Mainly for numerics
```
***

# Cleaning Data

***

Here are some of the problems we need to fix:

* In the "member_casual" column, there are two names for members ("member" and "Subscriber") and two names for casual riders ("Customer" and "casual"). We will need to consolidate that from four to two labels.

* The data can only be aggregated at the ride-level, which is too granular. We will want to add some additional columns of data -- such as day, month, year -- that provide additional opportunities to aggregate the data.

* We will want to add a calculated field for length of ride since the 2020Q1 data did not have the "tripduration" column. We will add "ride_length" to the entire dataframe for consistency.

* There are some rides where "tripduration" shows up as negative, including several hundred rides where Divvy took bikes out of circulation for Quality Control reasons. We will want to delete these rides.

Let's see how many different labels there were for the types of members till now.

```{r table of different user type labels}
table(all_trips$member_casual)
```

Let's consolidate this to only 2 names: Subscriber & Customer.

```{r consolidating labels to 2}
all_trips <-  all_trips %>% 
  mutate(member_casual = recode(member_casual
                           ,"Subscriber" = "member"
                           ,"Customer" = "casual"))

```

Let's check again

```{r label recheck}
table(all_trips$member_casual)
```
Let's create columns for date, month, and year of each ride.

```{r date, month, year columns}
all_trips$date <- as.Date(all_trips$started_at) #The default format is yyyy-mm-dd
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")

```

Let's add a ride length column and calculate it.

```{r ride length calculation}
all_trips$ride_length <- difftime(all_trips$ended_at, all_trips$started_at)
```

Now, let's inspect the structure once.

```{r str of df}
str(all_trips)
```

Let's convert ride_length to numeric data type.

```{r ride_length numeric}
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
```

The dataframe includes a few hundred entries when bikes were taken out of docks and checked for quality by Divvy, or ride_length was negative. We will create a new version of the dataframe (v2) since data is being removed.

```{r remove maintenance trips}
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length < 0),]
```

***

# Data Analysis

***

```{r descriptive analysis}
summary(all_trips_v2$ride_length)
```

Let's compare members and casual riders.

```{r compare member and casual ride length}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)
```

Average ride time by each day.

```{r avg ride-time/day}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

Let's fix the order of the days of the week.

```{r fix day order}
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```

```{r avg-time/day after fixing}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

Let's analyze ridership data by weekday.

```{r ridership weekday}
all_trips_v2 %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(member_type = member_casual, weekday) %>%  #groups by usertype and weekday
  summarise(number_of_rides = n()  #calculates the number of rides and average duration
            ,average_duration = mean(ride_length)) %>% 		# calculates the average duration
  arrange(weekday)  #sorting
```
```{r most used bike stations}
all_trips_v2 %>%
  group_by(member_type = member_casual, start_station_name) %>%
  summarise(number_of_rides = n(), average_ride_length = mean(ride_length)) %>%
  filter(number_of_rides > 10000, member_type == 'casual') %>%
  arrange(-number_of_rides) 
```

***

# Visualization

***

Let's visualize the number of rides by rider-type.

```{r, fig.align = "center", fig.cap = "Number of rides on each weekday, by user type"}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>%
  summarise(number_of_rides = n()/1000
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") + ylab("Number of rides (in thousands)")
```

Let's create a visualization for average duration.

```{r, fig.align = "center", fig.cap = "Average ride duration for each weekday"}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")
```
```{r, fig.align = "center", fig.cap = "Average ride duration by user type for each month"}
all_trips_v2 %>% 
  mutate(month_name = month(started_at, label = TRUE)) %>%
  group_by(member_casual, month_name) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, month_name)  %>% 
  ggplot(aes(x = month_name, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")
```

```{r, fig.align = "center", fig.cap = "Number of rides(in thousands) by user type for each month"}
all_trips_v2 %>% 
  mutate(month_name = month(started_at, label = TRUE)) %>%
  group_by(member_casual, month_name) %>% 
  summarise(number_of_rides = n()/1000) %>% 
  arrange(member_casual, month_name)  %>% 
  ggplot(aes(x = month_name, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") + ylab("Number of rides (in thousands)")
```

***

# Observations

* The casual riders' average trip duration is about 1.91 times that of the members'.

* Number of rides are highest during the summer months of July, August, and September, so we should plan on reducing downtime for maintenance and recharging.

* Average duration of rides was highest in the winter months of January and February.

* We can see the top 10 most used stations to start rides by casual users along with the average ride time and use that to make informed decisions about how many bikes to keep there. 

* It seems that casual members used bikes for longer duration, however, members had more trips.

***